---
title: "XP-GWAS_SIMULATION"
author: "Sam McNeill"
date: "2026-07-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Genome-wide XP-GWAS power simulation
Tests:
- effect sizes
- expanded-pool dilution
- number of causal SNPs
- core vs expanded pool designs


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

# 1. USER SETTINGS
```{r}
# Pool sizes
CORE_N_RES <- 11
CORE_N_SUS <- 11

EXP_N_RES <- 27
EXP_N_SUS <- 31

# Sequencing depth per pool
DEPTH <- 100

# Genome-wide SNP count
# Start with 50,000 for testing, then increase to 300,000 for final
M_SNPS <- 300000

# Number of simulation replicates
# Start with 50 or 100 for testing; use 500 for final
NSIM <- 500

# Effect sizes to test:
# True Res-Sus allele-frequency difference in the core pools
EFFECT_SIZES <- seq(0.05, 0.80, by = 0.05)

# Number of causal SNPs to test
CAUSAL_VALUES <- c(1, 2, 5)

# Expanded-pool dilution values
# 1.0 = expanded pools retain 100% of core signal
# 0.8 = expanded pools retain 80% of core signal
# 0.7 = expanded pools retain 70% of core signal
# 0.6 = expanded pools retain 60% of core signal
DILUTION_VALUES <- c(1.0, 0.8, 0.7, 0.6)

# Multiple testing correction
ALPHA <- 0.05
CORRECTION <- "BH"

set.seed(1234)
```

# 2. P-VALUE FUNCTION
Helper function to compare allele frequencies between Res and Sus pools at every SNP with biological pool sampling plus sequencing read sampling
```{r}
pooled_p_with_bio_variance <- function(x1, depth1, x2, depth2, n_res, n_sus) {
  
  p1 <- x1 / depth1
  p2 <- x2 / depth2
  
  p_pool <- (x1 + x2) / (depth1 + depth2)
  
  # Total variance includes biological pool sampling plus sequencing read sampling
  se <- sqrt(
    p_pool * (1 - p_pool) *
      (
        1 / (2 * n_res) +
        1 / (2 * n_sus) +
        1 / depth1 +
        1 / depth2
      )
  )
  
  z <- (p1 - p2) / se
  pval <- 2 * pnorm(-abs(z))
  
  pval[is.na(pval) | is.infinite(pval)] <- 1
  
  return(pval)
}
```

# 3. SIMULATE ONE GENOME-WIDE XP-GWAS

```{r}

simulate_genome <- function(
    n_res,
    n_sus,
    depth = 100,
    m_snps = 300000,
    n_causal = 1,
    delta = 0.40,
    alpha = 0.05,
    correction = "BH"
) {
  
  # Background allele frequencies for all SNPs
  # Avoid allele frequencies too close to 0 or 1
  p0 <- runif(m_snps, min = 0.05, max = 0.95)
  
  # Most SNPs are null: same allele frequency in Res and Sus
  p_res <- p0
  p_sus <- p0
  
  # Randomly choose causal SNPs
  causal_index <- sample(seq_len(m_snps), size = n_causal, replace = FALSE)
  is_causal <- rep(FALSE, m_snps)
  is_causal[causal_index] <- TRUE
  
  # Set causal SNP allele frequencies
  # delta = true Res-Sus allele-frequency difference
  p_res[is_causal] <- 0.5 + delta / 2
  p_sus[is_causal] <- 0.5 - delta / 2
  
  # Biological/accession sampling into pools
  # Each accession is diploid, so each contributes 2 allele copies
  res_alt_alleles <- rbinom(m_snps, size = 2 * n_res, prob = p_res)
  sus_alt_alleles <- rbinom(m_snps, size = 2 * n_sus, prob = p_sus)
  
  # Actual allele frequency in the constructed DNA pools
  res_pool_freq <- res_alt_alleles / (2 * n_res)
  sus_pool_freq <- sus_alt_alleles / (2 * n_sus)
  
  # Sequencing/read sampling
  res_alt_reads <- rbinom(m_snps, size = depth, prob = res_pool_freq)
  sus_alt_reads <- rbinom(m_snps, size = depth, prob = sus_pool_freq)
  
  # P-values for Res vs Sus at every SNP
  pval <- pooled_p_with_bio_variance(
    x1 = res_alt_reads,
    depth1 = depth,
    x2 = sus_alt_reads,
    depth2 = depth,
    n_res = n_res,
    n_sus = n_sus
  )
  
  # Multiple-testing correction
  adj_p <- p.adjust(pval, method = correction)
  significant <- adj_p < alpha
  
  # Causal SNP detection summaries
  n_causal_detected <- sum(significant & is_causal)
  
  causal_detected_any <- n_causal_detected >= 1
  causal_detected_half <- n_causal_detected >= ceiling(n_causal / 2)
  causal_detected_all <- n_causal_detected == n_causal
  
  # False positives among non-causal SNPs
  false_positives <- sum(significant & !is_causal)
  total_significant <- sum(significant)
  
  data.frame(
    n_res = n_res,
    n_sus = n_sus,
    depth = depth,
    m_snps = m_snps,
    n_causal = n_causal,
    delta = delta,
    alpha = alpha,
    correction = correction,
    n_causal_detected = n_causal_detected,
    causal_detected_any = causal_detected_any,
    causal_detected_half = causal_detected_half,
    causal_detected_all = causal_detected_all,
    false_positives = false_positives,
    total_significant = total_significant,
    min_causal_p = min(pval[is_causal]),
    min_causal_adj_p = min(adj_p[is_causal])
  )
}

```

# 4. ESTIMATE POWER ACROSS MANY SIMULATIONS

```{r}
estimate_genome_power <- function(
    n_res,
    n_sus,
    depth = 100,
    m_snps = 300000,
    n_causal = 1,
    delta = 0.40,
    alpha = 0.05,
    correction = "BH",
    nsim = 500
) {
  
  sims <- bind_rows(
    replicate(
      nsim,
      simulate_genome(
        n_res = n_res,
        n_sus = n_sus,
        depth = depth,
        m_snps = m_snps,
        n_causal = n_causal,
        delta = delta,
        alpha = alpha,
        correction = correction
      ),
      simplify = FALSE
    )
  )
  
  data.frame(
    n_res = n_res,
    n_sus = n_sus,
    depth = depth,
    m_snps = m_snps,
    n_causal = n_causal,
    delta = delta,
    alpha = alpha,
    correction = correction,
    nsim = nsim,
    
    # Power metrics
    power_any = mean(sims$causal_detected_any),
    beta_any = 1 - mean(sims$causal_detected_any),
    
    power_half = mean(sims$causal_detected_half),
    beta_half = 1 - mean(sims$causal_detected_half),
    
    power_all = mean(sims$causal_detected_all),
    beta_all = 1 - mean(sims$causal_detected_all),
    
    mean_causal_detected = mean(sims$n_causal_detected),
    
    # False-positive metrics
    mean_false_positives = mean(sims$false_positives),
    prob_any_false_positive = mean(sims$false_positives > 0),
    mean_total_significant = mean(sims$total_significant),
    
    # Useful diagnostic values
    mean_min_causal_p = mean(sims$min_causal_p),
    mean_min_causal_adj_p = mean(sims$min_causal_adj_p)
  )
}
```

# 5. RUN SIMULATIONS ACROSS EFFECT SIZES, CAUSAL SNP COUNTS, AND EXPANDED-POOL DILUTION VALUES
```{r}

system.time({

  simulation_results <- bind_rows(
    
    lapply(CAUSAL_VALUES, function(n_causal_value) {
      
      bind_rows(
        
        lapply(EFFECT_SIZES, function(delta_value) {
          
          # -----------------------------
          # Core pools
          # -----------------------------
          core_result <- estimate_genome_power(
            n_res = CORE_N_RES,
            n_sus = CORE_N_SUS,
            depth = DEPTH,
            m_snps = M_SNPS,
            n_causal = n_causal_value,
            delta = delta_value,
            alpha = ALPHA,
            correction = CORRECTION,
            nsim = NSIM
          ) %>%
            mutate(
              pool_design = "Core",
              design_label = paste0("Core: ", CORE_N_RES, " vs ", CORE_N_SUS),
              dilution_factor = 1.0,
              original_delta = delta_value,
              tested_delta = delta_value
            )
          
          # -----------------------------
          # Expanded pools under multiple dilution assumptions
          # -----------------------------
          expanded_results <- bind_rows(
            
            lapply(DILUTION_VALUES, function(dilution_value) {
              
              expanded_delta <- delta_value * dilution_value
              
              estimate_genome_power(
                n_res = EXP_N_RES,
                n_sus = EXP_N_SUS,
                depth = DEPTH,
                m_snps = M_SNPS,
                n_causal = n_causal_value,
                delta = expanded_delta,
                alpha = ALPHA,
                correction = CORRECTION,
                nsim = NSIM
              ) %>%
                mutate(
                  pool_design = "Expanded",
                  design_label = paste0(
                    "Expanded: ", EXP_N_RES, " vs ", EXP_N_SUS,
                    ", ", dilution_value * 100, "% signal retained"
                  ),
                  dilution_factor = dilution_value,
                  original_delta = delta_value,
                  tested_delta = expanded_delta
                )
            })
          )
          
          bind_rows(core_result, expanded_results)
        })
      )
    })
  )

Sys.sleep(2) 
})
```

We need to do this on HPC because compute time is too long. 

# 6. Read in CSV with results and format for plotting
```{r}
simulation_results <- read.csv("2000simulation_100depth_results_combined_1.csv")
```

```{r}

simulation_results <- simulation_results %>%
  mutate(
    design_label_clean = case_when(
      pool_design == "Core" ~ "Core 11 vs 11",
      pool_design == "Expanded" & dilution_factor == 1.0 ~ "Expanded 100%",
      pool_design == "Expanded" & dilution_factor == 0.88 ~ "Expanded 88%",
      pool_design == "Expanded" & dilution_factor == 0.46 ~ "Expanded 46%",
      pool_design == "Expanded" & dilution_factor == 0.2 ~ "Expanded 20%",
      TRUE ~ "Unknown"
    )
  )

simulation_results <- simulation_results %>%
  mutate(
    design_label_clean = factor(
      design_label_clean,
      levels = c(
        "Core 11 vs 11",
        "Expanded 100%",
        "Expanded 88%",
        "Expanded 46%",
        "Expanded 20%"
      )
    )
  )


simulation_results <- simulation_results %>%
  mutate(
    n_causal_label = factor(
      n_causal,
      levels = c(1, 2, 5),
      labels = c("1 causal SNP", "2 causal SNPs", "5 causal SNPs")
    )
  )

design_colors <- c(
  "Core 11 vs 11" = "black",
  "Expanded 100%" = "#009E73",
  "Expanded 88%" = "#0072B2",
  "Expanded 46%" = "#E69F00",
  "Expanded 20%" = "#D55E00"
)

# Quick look at results
simulation_results %>%
  select(
    pool_design,
    n_causal,
    original_delta,
    tested_delta,
    dilution_factor,
    power_any,
    power_half,
    power_all,
    mean_causal_detected,
    mean_false_positives,
    prob_any_false_positive
  )

```

# 7. PLOT: POWER TO DETECT AT LEAST ONE CAUSAL SNP

```{r}

ribbon_df <- simulation_results %>%
  filter(design_label_clean %in% c("Expanded 46%", "Expanded 88%")) %>%
  select(original_delta, n_causal_label, design_label_clean, power_any) %>%
  pivot_wider(
    names_from = design_label_clean,
    values_from = power_any
  )

p1 <- ggplot(
  simulation_results,
  aes(
    x = original_delta,
    y = power_any,
    color = design_label_clean
  )
) +
  geom_ribbon(
    data = ribbon_df,
    aes(
      x = original_delta,
      ymin = `Expanded 46%`,
      ymax = `Expanded 88%`
    ),
    inherit.aes = FALSE,
    fill = "gray70",
    alpha = 0.4
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ n_causal_label) +
  scale_color_manual(values = design_colors) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0.1, 0.8, by = 0.1)) +
  labs(
    title = "Fig 1. Power to Detect At Least One Causal SNP",
    subtitle = paste0(
      "300,000 SNPs, ",
      "100x depth, ",
      "2000 simulations, BH FDR < ", 0.05
      ),
      x = "Effect size: allele-frequency difference between Res and Sus pools",
      y = "Estimated power",
      color = "Design"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      legend.text = element_text(size = 12),
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )

p1
```

# 8. PLOT: POWER TO DETECT ALL CAUSAL SNPs

```{r}
p2 <- ggplot(
    simulation_results,
    aes(
      x = original_delta,
      y = power_all,
      color = design_label_clean
    )
  ) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 2) +
    facet_wrap(~ n_causal_label) +
    scale_color_manual(values = design_colors) +
    scale_y_continuous(limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0.1, 0.8, by = 0.1)) +
    labs(
      title = "Fig 2.Power to Detect All Causal SNPs",
      subtitle = paste0(
        "300,000 SNPs, ",
        "100x depth, ",
        "2000 simulations, BH FDR < ", 0.05
      ),
      x = "Effect size: allele-frequency difference between Res and Sus pools",
      y = "Estimated power",
      color = "Design"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      legend.text = element_text(size = 12),
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )

p2
```

# 9. PLOT: MEAN NUMBER OF CAUSAL SNPs DETECTED

```{r}
p3 <- ggplot(
  simulation_results,
  aes(
    x = original_delta,
    y = mean_causal_detected,
    color = design_label_clean
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ n_causal_label) +
  scale_color_manual(values = design_colors) +
  scale_x_continuous(breaks = seq(0.1, 0.8, by = 0.1)) +
  labs(
    title = "Fig 3.Mean Number of Causal SNPs Detected",
    subtitle = paste0(
        "300,000 SNPs, ",
        "100x depth, ",
        "2000 simulations, BH FDR < ", 0.05
    ),
    x = "Effect size: allele-frequency difference between Res and Sus pools",
    y = "Mean causal SNPs detected",
    color = "Design"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p3
```

# 10. PLOT: FALSE POSITIVES

```{r}
p4 <- ggplot(
  simulation_results,
  aes(
    x = original_delta,
    y = mean_false_positives,
    color = design_label_clean
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ n_causal_label) +
  scale_color_manual(values = design_colors) +
  scale_x_continuous(breaks = seq(0.1, 0.8, by = 0.1)) +
  labs(
    title = "Fig 4.Mean False Positives",
    subtitle = paste0(
        "300,000 SNPs, ",
        "100x depth, ",
        "2000 simulations, BH FDR < ", 0.05
    ),
    x = "Effect size: allele-frequency difference between Res and Sus pools",
    y = "Mean false positives",
    color = "Design"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p4
```

# Step 11: Save Plots
```{r}
ggsave("plot1_power_any.png", p1, width = 10, height = 6, dpi = 300)
ggsave("plot2_mean_causal_detected.png", p2, width = 10, height = 6, dpi = 300)
ggsave("plot3_power_all.png", p3, width = 10, height = 6, dpi = 300)
ggsave("plot4_false_positives.png", p4, width = 10, height = 6, dpi = 300)


pdf("all_xpgwas_power_plots.pdf", width = 10, height = 6)

print(p1)
print(p2)
print(p3)
print(p4)

dev.off()
```
